Explore the SingleR results

### 2025-07-15 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
got dotplots working with new marker genes and labels, including use of updated palettes

# palette file
palette_file <- file.path(
  module_dir,
  "palettes",
  "harmonized-cell-type-palette.tsv"
)

@@

# set up palette for umaps
palette_df <- readr::read_tsv(palette_file)
celltype_pal <- palette_df$hex_color
names(celltype_pal) <- palette_df$harmonized_label
```

@@

```{r, fig.width = 14, fig.height = 7}
ggplot(celltype_recoded_df) +
  aes(x = UMAP1, y = UMAP2, color = label_recoded) +
  geom_point(size = 0.25, alpha = 0.5) +
  scale_color_manual(
    values = celltype_pal,
    name = "Harmonized cell types"
  ) +
  facet_wrap(vars(annotation_type)) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

@@

```{r fig.width = 16, fig.height = 16}
celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler") |>
  faceted_umap(
    annotation_column = label_recoded,
    celltype_colors = celltype_pal
  )
```

@@

@@

faceted_umap(
  celltype_facet_df,
  annotation_column = label_recoded,
  celltype_colors = celltype_pal,
  facet_type = "grid",
  annotation_type_column = annotation_type
)
```

@@

```{r}
# Prepare the expressed_genes vector
# we only care about if that gene is expressed otherwise we won't waste memory and include it
gene_sums <- rowData(merged_sce) |>
  as.data.frame() |>
  dplyr::select(contains("detected")) |>
  as.matrix() |>
  rowSums()
expressed_genes <- names(gene_sums)[gene_sums > 0]

# Prepare data frame with singler labels to plot
singler_recoded_df <- celltype_recoded_df |>
  dplyr::filter(
    annotation_type == "singler",
    # we don't consider the NAs here
    label_recoded != "unknown_singler"
  ) |>
  dplyr::select(cell_id, label_recoded)

@@

nbatlas_markers_df <- nbatlas_markers_df |>
  # keep only the top `n_marker_genes`
  dplyr::filter(direction == "up") |>
  dplyr::group_by(NBAtlas_label) |>
  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
  dplyr::ungroup() |>
  dplyr::filter(rank_lfc <= n_marker_genes) |>
  harmonize_celltypes(NBAtlas_label, marker_gene_label) |>
  dplyr::select(marker_gene_label, gene_symbol, ensembl_gene_id)

nbatlas_bar_order <- total_cells_df$label_recoded
```

```{r, fig.width = 24, fig.height = 12}
generate_dotplot(
  merged_sce,
  nbatlas_markers_df,
  singler_recoded_df,
  total_cells_df,
  expressed_genes,
  nbatlas_bar_order
)
```

@@

```{r}
# prepare for dotplot

consensus_markers_df <- consensus_markers_df |>
  # we'll use this to help determine the bar order
  harmonize_celltypes(NBAtlas_label, NBAtlas_label_recoded) |>
  dplyr::select(
    marker_gene_label = validation_group_annotation,
    ensembl_gene_id,
    gene_symbol,
    NBAtlas_label_recoded
  )

# get the bar order
consensus_bar_order <- total_cells_df |>
  dplyr::select(label_recoded, y_label) |>
  # inner!!!!
  dplyr::inner_join(consensus_markers_df, by = c("label_recoded" = "NBAtlas_label_recoded")) |>
  dplyr::arrange(y_label) |>
  dplyr::pull(marker_gene_label) |>
  unique()
```

```{r, fig.width = 22, fig.height = 12}
generate_dotplot(
  merged_sce,
  consensus_markers_df,
  singler_recoded_df,
  total_cells_df,
  expressed_genes,
  consensus_bar_order
)
```

>>>>

# palette files
recoded_palette_file <- file.path(
  module_dir,
  "palettes",
  "harmonized-cell-type-palette.tsv"
)

nbatlas_palette_file <- file.path(
  module_dir,
  "palettes",
  "nbatlas-marker-genes-palette.tsv"
)

@@

# set up palettes

# recoded to shared colors
recoded_palette_df <- readr::read_tsv(recoded_palette_file)
recoded_celltype_pal <- recoded_palette_df$hex_color
names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label

# only the NBAtlas labels - use for validation dotplot
nbatlas_palette_df <- readr::read_tsv(nbatlas_palette_file)
nbatlas_celltype_pal <- nbatlas_palette_df$hex_color
names(nbatlas_celltype_pal) <- nbatlas_palette_df$NBAtlas_label

@@

```{r, fig.width = 14, fig.height = 7}
ggplot(celltype_recoded_df) +
  aes(x = UMAP1, y = UMAP2, color = label_recoded) +
  geom_point(size = 0.25, alpha = 0.5) +
  scale_color_manual(
    values = recoded_celltype_pal,
    name = "Harmonized cell types"
  ) +
  facet_wrap(vars(annotation_type)) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

@@

```{r fig.width = 16, fig.height = 16}
celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler") |>
  faceted_umap(
    annotation_column = label_recoded,
    celltype_colors = recoded_celltype_pal
  )
```

@@

Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid cagetory contains additonal cell types.

@@

faceted_umap(
  celltype_facet_df,
  annotation_column = label_recoded,
  celltype_colors = recoded_celltype_pal,
  facet_type = "grid",
  annotation_type_column = annotation_type
)
```

@@

```{r}
# Prepare the expressed_genes vector
# we only care about if that gene is expressed otherwise we won't waste memory and include it
gene_sums <- rowData(merged_sce) |>
  as.data.frame() |>
  dplyr::select(contains("detected")) |>
  as.matrix() |>
  rowSums()
expressed_genes <- names(gene_sums)[gene_sums > 0]
```

@@

top_nbatlas_markers_df <- nbatlas_markers_df |>
  # keep only the top `n_marker_genes`
  dplyr::filter(direction == "up") |>
  dplyr::group_by(NBAtlas_label) |>
  dplyr::mutate(rank_lfc = rank(-avg_log2FC)) |>
  dplyr::ungroup() |>
  dplyr::filter(rank_lfc <= n_marker_genes) |>
  dplyr::select(marker_gene_label = NBAtlas_label, gene_symbol, ensembl_gene_id)
```

```{r, fig.width = 24, fig.height = 12}
# we want to plot the literal labels here
full_celltype_df <- celltype_df |>
  dplyr::select(
    cell_id, 
    # the dotplot function expects this column name
    label_recoded = pruned.labels
  ) |>
  dplyr::mutate(
    label_recoded = dplyr::case_when(
      # collapse the dendritic cells & monocytes to match what's in the marker genes
      stringr::str_detect(label_recoded, "cDC") ~ "cDC", 
      stringr::str_detect(label_recoded, "monocyte") ~ "Monocyte",
      label_recoded == "NE" ~ "Neuroendocrine",
      .default = label_recoded
  )) |>
  dplyr::filter(label_recoded != "unknown_singler")

# get total number of cells per final annotation group and set up y_label
total_cells_df <- full_celltype_df |>
  dplyr::count(label_recoded, name = "total_cells") |>
  dplyr::arrange(desc(total_cells)) |>
  dplyr::mutate(y_label = glue::glue("{label_recoded} ({total_cells})"))

singler_order <- total_cells_df$y_label
total_cells_df$y_label <- factor(total_cells_df$y_label, levels = singler_order)
nbatlas_bar_order <- total_cells_df$label_recoded
```

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  singler_df = full_celltype_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```

@@

```{r}
# Prepare data frame with singler labels to plot
singler_recoded_df <- celltype_recoded_wide_df |>
  # we don't consider the NAs here
  dplyr::filter(singler != "unknown_singler") |>
  dplyr::select(cell_id, label_recoded = singler)

```{r}
# prepare markers for dotplot
dotplot_consensus_markers_df <- consensus_markers_df |>
  # consider only matching labels
  dplyr::filter(validation_group_annotation %in% singler_recoded_df$label_recoded) |>
  dplyr::select(
    marker_gene_label = validation_group_annotation,
    ensembl_gene_id,
    gene_symbol
  )

# get the bar order
consensus_bar_order <- total_cells_df |>
  dplyr::select(label_recoded, y_label) |>
  # inner!!!!
  dplyr::inner_join(
    dotplot_consensus_markers_df, 
    by = c("label_recoded" = "marker_gene_label")
  ) |>
  dplyr::arrange(y_label) |>
  dplyr::pull(label_recoded) |>
  unique()
```

```{r, fig.width = 26, fig.height = 12}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = dotplot_consensus_markers_df,
  singler_df = singler_recoded_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = consensus_bar_order, 
  celltype_palette = recoded_celltype_pal, 
  min_cells = 0
)
```
